peter_models

### 2024-12-08 peter_models.qmd
updating interactions model
updating modeling script

```{r}
# Create a summary table for all categorical variables
variables <- c("Q29", "Q271", "S2_RACE_PRIME", "S3", "S3B", "S5_AGE", "S13", "S14", "S7", "Q21", "S2_M", "Q627", "Q628", "Q629R1", "Religion", "educ", "sex_orientation", "rural", "white", "race_recoded", "age_recoded")

# Chi-square tests for each variable
chi_square_results <- lapply(variables, function(var) {
  chisq.test(table(dataset_clean[[var]], dataset_clean$Q29))
})

# Print chi-square results
names(chi_square_results) <- variables
print(chi_square_results)
```

@@

```{r}
# generate predictions for the ordinal model

new_dataset <- data.frame

predict(ord_mod, new_dataset)
```

@@

ord_mod_interactions <- polr(Q29 ~ Q271_reordered + race_recoded +
                  age_recoded +
            educ +  # Educational Attainment
            S3B + # Gender
            Q21, # Party affiliation
            # Q271_reordered * race_recoded,  # Interaction term
          data = dataset_clean, Hess=TRUE)

@@

new_dataset <- data.frame

predict(ord_mod, new_dataset)
```

@@

# Load packages
library(nnet)  # for multinomial logistic regression
library(tidyverse)
library(car)
library(effects)

# ensure outcome variable is a factor but not ordered
dataset_clean$Q29_unordered <- factor(dataset_clean$Q29, ordered = FALSE)
dataset_clean$Q271_simple <- as.numeric(dataset_clean$Q271)

# Fit multinomial model
m_mult <- multinom(Q29_unordered ~ Q271 + race_recoded + age_recoded +
            # rural + # Rural or Urban/Suburban
            # sex_orientation + # Sexual Orientation
            educ + # Educational Attainment
            S3B + # Gender
            # S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean)

# Calculate predicted probabilities of marginal effects
# Create sequence of Q271_simple values
x_vals <- seq(min(dataset_clean$Q271_simple), max(dataset_clean$Q271_simple), length.out = 100)

# Function to calculate average predicted probabilities
get_avg_probs <- function(model, newdata, var_name) {
    preds <- predict(model, newdata = newdata, type = "probs")
    colMeans(preds)
}

# Calculate marginal effects by race
races <- unique(dataset_clean$race_recoded)
marg_effects <- list()

for(race in races) {
    probs <- matrix(0, nrow = length(x_vals), ncol = length(levels(dataset_clean$Q29_unordered)))
    for(i in seq_along(x_vals)) {
        temp_data <- dataset_clean
        temp_data$Q271_simple <- x_vals[i]
        temp_data$race_major <- race
        probs[i,] <- get_avg_probs(m_mult, temp_data, "Q271_simple")
    }
    marg_effects[[race]] <- data.frame(
        x = x_vals,
        race = race,
        prob1 = probs[,1],
        prob2 = probs[,2],
        prob3 = probs[,3],
        prob4 = probs[,4]
    )
}

# Combine all marginal effects
marg_effects_df <- bind_rows(marg_effects)

# Create plots for marginal effects
# Reshape data for plotting
marg_effects_long <- marg_effects_df %>%
    pivot_longer(cols = starts_with("prob"),
                names_to = "outcome",
                values_to = "probability") %>%
    mutate(outcome = factor(outcome,
                          levels = paste0("prob", 1:4),
                          labels = levels(dataset_clean$Q29_unordered)))

# Plot marginal effects
p1 <- ggplot(marg_effects_long, 
             aes(x = x, y = probability, color = outcome)) +
    geom_line() +
    facet_wrap(~race) +
    theme_minimal() +
    labs(title = "Marginal Effects of Racial Identity Importance by Race",
         x = "Importance of Racial Identity",
         y = "Predicted Probability",
         color = "Political Interest Level") +
    theme(legend.position = "bottom")
print(p1)

# Model fit statistics and comparisons
print("Model summaries:")
print(summary(m_mult))

# Calculate McFadden's R-squared
null_model <- multinom(Q29_unordered ~ 1, data = dataset_clean)
mcfadden_r2 <- 1 - logLik(m_mult)/logLik(null_model)

print("McFadden's R-squared:")
print(mcfadden_r2)

# Classification accuracy
pred_class <- predict(m_mult)
conf_matrix <- table(pred_class, dataset_clean$Q29_unordered)
accuracy <- sum(diag(conf_matrix))/sum(conf_matrix)

print("Classification accuracy:")
print(accuracy)
print("Confusion Matrix:")
print(conf_matrix)

# Print AIC
print(AIC(m_mult))

# Calculate predicted probabilities
pred_data <- expand.grid(
    Q271 = unique(dataset_clean$Q271),
    age_recoded = unique(dataset_clean$age_recoded),
    race_recoded = unique(dataset_clean$race_recoded),
    educ = unique(dataset_clean$educ),
    S3B = unique(dataset_clean$S3B),
    Q21 = unique(dataset_clean$Q21)
)

# Get predictions
preds <- predict(m_mult, newdata = pred_data, type = "probs")
pred_data <- cbind(pred_data, as.data.frame(preds))

# Create separate plots for each outcome level
pred_long <- pred_data %>%
    pivot_longer(cols = starts_with("("),
                names_to = "interest_level",
                values_to = "probability")

# Plot for each outcome level
p2 <- ggplot(pred_long, 
             aes(x = Q271, y = probability, color = race_recoded)) +
    geom_line() +
    facet_wrap(~interest_level) +
    theme_minimal() +
    labs(title = "Predicted Probabilities by Outcome Level",
         x = "Importance of Racial Identity",
         y = "Probability",
         color = "Race/Ethnicity") +
    theme(legend.position = "bottom")
print(p2)

@@

```{r}
dataset_clean$Q271_simple <- as.numeric(dataset_clean$Q271)
# dataset_clean$Q271 <- factor(dataset_clean$Q271,
#                             levels = c("(5) Not at all important",
#                                      "(4) Slightly important",
#                                      "(3) Moderately important",
#                                      "(2) Very important",
#                                      "(1) Extremely important"),
#                             ordered = TRUE)

m1 <- clm(Q29 ~ Q271 + race_recoded + age_recoded +
            rural + # Rural or Urban/Suburban
            sex_orientation + # Sexual Orientation
            Religion + 
            educ + # Educational Attainment
            S3B + # Gender
            S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean)

summary(m1)

# Check for multicollinearity using a linear approximation
vif_m1 <- vif(lm(as.numeric(Q29) ~ Q271 + race_recoded + age_recoded +
            rural + # Rural or Urban/Suburban
            sex_orientation + # Sexual Orientation
            Religion + 
            educ + # Educational Attainment
            S3B + # Gender
            S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean))

print("Variance Inflation Factors:")
print(vif_m1)

# Test proportional odds assumption
print("Testing proportional odds assumption:")
test1 <- nominal_test(m1)
print("Nominal test for Model 1:")
print(test1)

# Alternative diagnostic approach using linear approximation
lm_approx <- lm(as.numeric(Q29) ~ Q271 + race_recoded + age_recoded +
            rural + # Rural or Urban/Suburban
            sex_orientation + # Sexual Orientation
            Religion + 
            educ + # Educational Attainment
            S3B + # Gender
            S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean)
diagnostics_df <- data.frame(
  fitted = fitted(lm_approx),
  residuals = residuals(lm_approx),
  index = 1:nrow(dataset_clean)
)

# Plot residuals vs fitted values using linear approximation
p1 <- ggplot(diagnostics_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "loess") +
  theme_minimal() +
  labs(title = "Residuals vs Fitted Values (Linear Approximation)",
       x = "Fitted Values",
       y = "Residuals")
print(p1)

# Calculate and plot Cook's distance
cooks_d <- cooks.distance(lm_approx)
influence_df <- data.frame(
  index = 1:length(cooks_d),
  cooks_distance = cooks_d
)

p2 <- ggplot(influence_df, aes(x = index, y = cooks_distance)) +
  geom_point() +
  geom_hline(yintercept = 4/nrow(dataset_clean), color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Cook's Distance Plot",
       x = "Observation Number",
       y = "Cook's Distance")
print(p2)

# Count influential observations
influential_obs <- which(cooks_d > 4/nrow(dataset_clean))
print("Number of influential observations:")
print(length(influential_obs))

# Create prediction visualization
pred_data <- expand.grid(
  Q271 = levels(dataset_clean$Q271),
  race_recoded = levels(dataset_clean$race_recoded), 
  age_recoded = levels(dataset_clean$age_recoded),
  rural = levels(dataset_clean$rural),
  sex_orientation = levels(dataset_clean$sex_orientation),
  Religion = levels(dataset_clean$Religion),
  educ = levels(dataset_clean$educ),
  S3B = levels(dataset_clean$S3B),
  S7 = levels(dataset_clean$S7),
  Q21 = levels(dataset_clean$Q21)
)

# Get predicted probabilities
preds <- predict(m1, newdata = pred_data, type = "prob")
# Convert predictions to numeric scores (weighted average)
pred_data$predicted_score <- apply(preds$fit, 1, function(x) sum(x * 1:4))

# Create prediction plot
p3 <- ggplot(pred_data, aes(x = Q271, y = predicted_score, color = race_recoded)) +
  geom_point() +
  geom_line(aes(group = race_recoded)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Predicted Political Interest by Racial Identity Importance and Race",
       x = "Importance of Racial Identity",
       y = "Predicted Political Interest Level (1-4)",
       color = "Race/Ethnicity")
print(p3)

# Print the coefficients table
print("Coefficients for Model 1:")
coef_table <- summary(m1)$coefficients
print(coef_table)

# Calculate and print odds ratios for the coefficients
odds_ratios <- exp(coef_table[, "Estimate"])
print("Odds Ratios:")
print(odds_ratios)

>>>>

```{r}
# # Create a summary table for all categorical variables
# variables <- c("Q29", "Q271", "S2_RACE_PRIME", "S3", "S3B", "S5_AGE", "S13", "S14", "S7", "Q21", "S2_M", "Q627", "Q628", "Q629R1", "Religion", "educ", "sex_orientation", "rural", "white", "race_recoded", "age_recoded")
# 
# 
# # Chi-square tests for each variable
# chi_square_results <- lapply(variables, function(var) {
#   chisq.test(table(dataset_clean[[var]], dataset_clean$Q29))
# })
# 
# # Print chi-square results
# names(chi_square_results) <- variables
# print(chi_square_results)
```

@@

```{r}
# generate predictions for the ordinal model
# 
new_dataset <- data.frame

# predict(ord_mod, new_dataset)
```

@@

ord_mod_interactions <- polr(Q29 ~ Q271_reordered * S3B +
                               race_recoded +
                  age_recoded +
            educ +  # Educational Attainment
            Q21, # Party affiliation
          data = dataset_clean, Hess=TRUE)

@@

# new_dataset <- data.frame
# 
# predict(ord_mod, new_dataset)
```

@@

# 
# # Load packages
# library(nnet)  # for multinomial logistic regression
# library(tidyverse)
# library(car)
# library(effects)
# 
# 
# # ensure outcome variable is a factor but not ordered
# dataset_clean$Q29_unordered <- factor(dataset_clean$Q29, ordered = FALSE)
# dataset_clean$Q271_simple <- as.numeric(dataset_clean$Q271)
# 
# # Fit multinomial model
# m_mult <- multinom(Q29_unordered ~ Q271 + race_recoded + age_recoded +
#             # rural + # Rural or Urban/Suburban
#             # sex_orientation + # Sexual Orientation
#             educ + # Educational Attainment
#             S3B + # Gender
#             # S7 + # Nativity
#             Q21, # Party affiliation
#           data = dataset_clean)
# 
# 
# 
# # Calculate predicted probabilities of marginal effects
# # Create sequence of Q271_simple values
# x_vals <- seq(min(dataset_clean$Q271_simple), max(dataset_clean$Q271_simple), length.out = 100)
# 
# # Function to calculate average predicted probabilities
# get_avg_probs <- function(model, newdata, var_name) {
#     preds <- predict(model, newdata = newdata, type = "probs")
#     colMeans(preds)
# }
# 
# # Calculate marginal effects by race
# races <- unique(dataset_clean$race_recoded)
# marg_effects <- list()
# 
# for(race in races) {
#     probs <- matrix(0, nrow = length(x_vals), ncol = length(levels(dataset_clean$Q29_unordered)))
#     for(i in seq_along(x_vals)) {
#         temp_data <- dataset_clean
#         temp_data$Q271_simple <- x_vals[i]
#         temp_data$race_major <- race
#         probs[i,] <- get_avg_probs(m_mult, temp_data, "Q271_simple")
#     }
#     marg_effects[[race]] <- data.frame(
#         x = x_vals,
#         race = race,
#         prob1 = probs[,1],
#         prob2 = probs[,2],
#         prob3 = probs[,3],
#         prob4 = probs[,4]
#     )
# }
# 
# # Combine all marginal effects
# marg_effects_df <- bind_rows(marg_effects)
# 
# # Create plots for marginal effects
# # Reshape data for plotting
# marg_effects_long <- marg_effects_df %>%
#     pivot_longer(cols = starts_with("prob"),
#                 names_to = "outcome",
#                 values_to = "probability") %>%
#     mutate(outcome = factor(outcome,
#                           levels = paste0("prob", 1:4),
#                           labels = levels(dataset_clean$Q29_unordered)))
# 
# # Plot marginal effects
# p1 <- ggplot(marg_effects_long, 
#              aes(x = x, y = probability, color = outcome)) +
#     geom_line() +
#     facet_wrap(~race) +
#     theme_minimal() +
#     labs(title = "Marginal Effects of Racial Identity Importance by Race",
#          x = "Importance of Racial Identity",
#          y = "Predicted Probability",
#          color = "Political Interest Level") +
#     theme(legend.position = "bottom")
# print(p1)
# 
# # Model fit statistics and comparisons
# print("Model summaries:")
# print(summary(m_mult))
# 
# # Calculate McFadden's R-squared
# null_model <- multinom(Q29_unordered ~ 1, data = dataset_clean)
# mcfadden_r2 <- 1 - logLik(m_mult)/logLik(null_model)
# 
# print("McFadden's R-squared:")
# print(mcfadden_r2)
# 
# # Classification accuracy
# pred_class <- predict(m_mult)
# conf_matrix <- table(pred_class, dataset_clean$Q29_unordered)
# accuracy <- sum(diag(conf_matrix))/sum(conf_matrix)
# 
# print("Classification accuracy:")
# print(accuracy)
# print("Confusion Matrix:")
# print(conf_matrix)
# 
# # Print AIC
# print(AIC(m_mult))
# 
# # Calculate predicted probabilities
# pred_data <- expand.grid(
#     Q271 = unique(dataset_clean$Q271),
#     age_recoded = unique(dataset_clean$age_recoded),
#     race_recoded = unique(dataset_clean$race_recoded),
#     educ = unique(dataset_clean$educ),
#     S3B = unique(dataset_clean$S3B),
#     Q21 = unique(dataset_clean$Q21)
# )
# 
# # Get predictions
# preds <- predict(m_mult, newdata = pred_data, type = "probs")
# pred_data <- cbind(pred_data, as.data.frame(preds))
# 
# # Create separate plots for each outcome level
# pred_long <- pred_data %>%
#     pivot_longer(cols = starts_with("("),
#                 names_to = "interest_level",
#                 values_to = "probability")
# 
# 
# # Plot for each outcome level
# p2 <- ggplot(pred_long, 
#              aes(x = Q271, y = probability, color = race_recoded)) +
#     geom_line() +
#     facet_wrap(~interest_level) +
#     theme_minimal() +
#     labs(title = "Predicted Probabilities by Outcome Level",
#          x = "Importance of Racial Identity",
#          y = "Probability",
#          color = "Race/Ethnicity") +
#     theme(legend.position = "bottom")
# print(p2)

@@

# dataset_clean$Q271_simple <- as.numeric(dataset_clean$Q271)
# # dataset_clean$Q271 <- factor(dataset_clean$Q271,
# #                             levels = c("(5) Not at all important",
# #                                      "(4) Slightly important",
# #                                      "(3) Moderately important",
# #                                      "(2) Very important",
# #                                      "(1) Extremely important"),
# #                             ordered = TRUE)
# 
# m1 <- clm(Q29 ~ Q271 + race_recoded + age_recoded +
#             rural + # Rural or Urban/Suburban
#             sex_orientation + # Sexual Orientation
#             Religion + 
#             educ + # Educational Attainment
#             S3B + # Gender
#             S7 + # Nativity
#             Q21, # Party affiliation
#           data = dataset_clean)
# 
# summary(m1)
# 
# # Check for multicollinearity using a linear approximation
# vif_m1 <- vif(lm(as.numeric(Q29) ~ Q271 + race_recoded + age_recoded +
#             rural + # Rural or Urban/Suburban
#             sex_orientation + # Sexual Orientation
#             Religion + 
#             educ + # Educational Attainment
#             S3B + # Gender
#             S7 + # Nativity
#             Q21, # Party affiliation
#           data = dataset_clean))
# 
# print("Variance Inflation Factors:")
# print(vif_m1)
# 
# # Test proportional odds assumption
# print("Testing proportional odds assumption:")
# test1 <- nominal_test(m1)
# print("Nominal test for Model 1:")
# print(test1)
# 
# # Alternative diagnostic approach using linear approximation
# lm_approx <- lm(as.numeric(Q29) ~ Q271 + race_recoded + age_recoded +
#             rural + # Rural or Urban/Suburban
#             sex_orientation + # Sexual Orientation
#             Religion + 
#             educ + # Educational Attainment
#             S3B + # Gender
#             S7 + # Nativity
#             Q21, # Party affiliation
#           data = dataset_clean)
# diagnostics_df <- data.frame(
#   fitted = fitted(lm_approx),
#   residuals = residuals(lm_approx),
#   index = 1:nrow(dataset_clean)
# )
# 
# # Plot residuals vs fitted values using linear approximation
# p1 <- ggplot(diagnostics_df, aes(x = fitted, y = residuals)) +
#   geom_point(alpha = 0.1) +
#   geom_smooth(method = "loess") +
#   theme_minimal() +
#   labs(title = "Residuals vs Fitted Values (Linear Approximation)",
#        x = "Fitted Values",
#        y = "Residuals")
# print(p1)
# 
# # Calculate and plot Cook's distance
# cooks_d <- cooks.distance(lm_approx)
# influence_df <- data.frame(
#   index = 1:length(cooks_d),
#   cooks_distance = cooks_d
# )
# 
# p2 <- ggplot(influence_df, aes(x = index, y = cooks_distance)) +
#   geom_point() +
#   geom_hline(yintercept = 4/nrow(dataset_clean), color = "red", linetype = "dashed") +
#   theme_minimal() +
#   labs(title = "Cook's Distance Plot",
#        x = "Observation Number",
#        y = "Cook's Distance")
# print(p2)
# 
# # Count influential observations
# influential_obs <- which(cooks_d > 4/nrow(dataset_clean))
# print("Number of influential observations:")
# print(length(influential_obs))
# 
# 
# 
# # Create prediction visualization
# pred_data <- expand.grid(
#   Q271 = levels(dataset_clean$Q271),
#   race_recoded = levels(dataset_clean$race_recoded), 
#   age_recoded = levels(dataset_clean$age_recoded),
#   rural = levels(dataset_clean$rural),
#   sex_orientation = levels(dataset_clean$sex_orientation),
#   Religion = levels(dataset_clean$Religion),
#   educ = levels(dataset_clean$educ),
#   S3B = levels(dataset_clean$S3B),
#   S7 = levels(dataset_clean$S7),
#   Q21 = levels(dataset_clean$Q21)
# )
# 
# # Get predicted probabilities
# preds <- predict(m1, newdata = pred_data, type = "prob")
# # Convert predictions to numeric scores (weighted average)
# pred_data$predicted_score <- apply(preds$fit, 1, function(x) sum(x * 1:4))
# 
# # Create prediction plot
# p3 <- ggplot(pred_data, aes(x = Q271, y = predicted_score, color = race_recoded)) +
#   geom_point() +
#   geom_line(aes(group = race_recoded)) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   labs(title = "Predicted Political Interest by Racial Identity Importance and Race",
#        x = "Importance of Racial Identity",
#        y = "Predicted Political Interest Level (1-4)",
#        color = "Race/Ethnicity")
# print(p3)
# 
# # Print the coefficients table
# print("Coefficients for Model 1:")
# coef_table <- summary(m1)$coefficients
# print(coef_table)
# 
# # Calculate and print odds ratios for the coefficients
# odds_ratios <- exp(coef_table[, "Estimate"])
# print("Odds Ratios:")
# print(odds_ratios)
